---
title: "Estimation error vs trend parameter (R2.3)"
output:
  pdf_document:
    fig_height: 5
    fig_width: 10
    toc: yes
  html_notebook:
    code_folding: hide
---

Supplementary analyses addressing Reviewer 2, comment R2.3: estimation error as
a function of the trend parameter.

Uses the existing MCMC parameter traces from the calibration study (500 datasets).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
# Add user library for R 4.4 packages if running under R 4.5
user_lib <- paste0(Sys.getenv("HOME"), "/Library/R/arm64/4.4/library")
if (dir.exists(user_lib)) .libPaths(c(user_lib, .libPaths()))
```

```{r}
library(ape)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(patchwork)

theme_set(theme_bw(base_size = 12))

# Extant tips have labels starting with "t", fossils with "f"
getExtant <- function(tree) tree$tip.label[grepl("^t", tree$tip.label)]
```

# Load calibration data

```{r}
bioclust_path <- paste0(Sys.getenv("HOME"), "/bioclust/INSANE_OBD/Calibration/")
outputs_path  <- paste0(bioclust_path, "outputs/")
datasets_path <- paste0(bioclust_path, "datasets_calibration/")

dataset_nbs <- sort(as.integer(gsub("OBDD_SimulatedOBDD_dataset|_2000000iter.log", "",
  list.files(outputs_path, pattern = "OBDD_SimulatedOBDD_dataset.*2000000iter.log"))))

cat("Number of datasets:", length(dataset_nbs), "\n")

parameter_labels <- c(
  "λ"  = expression(lambda[0]),
  "μ"  = expression(mu[0]),
  "αλ" = expression(alpha[lambda]),
  "αμ" = expression(alpha[mu]),
  "σλ" = expression(sigma[lambda]),
  "σμ" = expression(sigma[mu]),
  "ψ1" = expression(psi[1]),
  "ψ2" = expression(psi[2]),
  "ω1" = expression(omega[1]),
  "ω2" = expression(omega[2])
)
```

```{r}
# Column name mapping for trace files
col_rename <- c(lambda_root = "λ", mu_root = "μ", alpha_lambda = "αλ", alpha_mu = "αμ",
                sigma_lambda = "σλ", sigma_mu = "σμ", psi_1 = "ψ1", psi_2 = "ψ2",
                omega_1 = "ω1", omega_2 = "ω2")
trace_cols <- names(col_rename)

parameters_OBDD <- do.call(rbind, lapply(dataset_nbs, function(dataset_nb) {
  # Read only the 10 parameter columns from the trace file
  trace_file <- paste0(outputs_path, "OBDD_SimulatedOBDD_dataset", dataset_nb, "_2000000iter.log")
  trace <- data.table::fread(trace_file, select = trace_cols, data.table = FALSE)
  names(trace) <- col_rename[names(trace)]

  # Read true parameter values
  params <- read.delim(paste0(datasets_path, "dataset", dataset_nb, "/params.csv")) %>%
    gather(par, value) %>% slice(1:10)

  # Tree info
  tree_rec <- read.tree(paste0(datasets_path, "dataset", dataset_nb, "/tree_reconstructed.tre"))
  ntips_rec <- Ntip(tree_rec)
  ntips_ext <- sum(grepl("^t", tree_rec$tip.label))
  nfossils  <- ntips_rec - ntips_ext

  # Count occurrences without reading the full file
  occ_file <- paste0(datasets_path, "dataset", dataset_nb, "/occurrences.csv")
  nocc <- ifelse(file.info(occ_file)$size > 0,
                 length(readLines(occ_file)), 0)

  data.frame(
    dataset_nb  = dataset_nb,
    par         = params$par,
    true_value  = params$value,
    median_post = sapply(trace[params$par], median),
    NMSE        = colMeans(t(t(trace[params$par]) - params$value)^2) / params$value^2,
    ntips       = ntips_rec,
    ntipsextant = ntips_ext,
    nfossils    = nfossils,
    nocc        = nocc,
    row.names   = NULL
  )
}))

parameters_OBDD$ParametersExpr <- factor(parameters_OBDD$par,
  levels = names(parameter_labels), labels = parameter_labels)
```


# R2.3: Estimation error vs trend parameter

The reviewer predicts that in models with a positive net-diversification trend
($\alpha_\lambda > 0$ or $\alpha_\mu < 0$), the estimation error should be smaller
because the tree is expected to be larger and more informative.

```{r}
# Join trend parameters to estimation errors
trend_data <- parameters_OBDD %>%
  left_join(
    parameters_OBDD %>%
      filter(par == "αλ") %>%
      select(dataset_nb, true_αλ = true_value),
    by = "dataset_nb"
  ) %>%
  left_join(
    parameters_OBDD %>%
      filter(par == "αμ") %>%
      select(dataset_nb, true_αμ = true_value),
    by = "dataset_nb"
  ) %>%
  mutate(net_div_trend = true_αλ - true_αμ)
```

## NMSE vs speciation trend $\alpha_\lambda$

```{r fig.width=10, fig.height=5}
p <- ggplot(trend_data %>% filter(!par %in% c("αλ", "αμ")),
       aes(x = true_αλ, y = NMSE)) +
  geom_point(alpha = 0.15, size = 1, color = "darkblue") +
  geom_smooth(method = "loess", color = "dodgerblue3", linewidth = 0.5, se = TRUE,
              fill = "dodgerblue3") +
  scale_y_log10(name = "Normalized Mean Squared Error") +
  scale_x_continuous(name = expression("True " * alpha[lambda] * " (speciation trend)")) +
  facet_wrap(~ ParametersExpr, scales = "fixed", nrow = 2, labeller = label_parsed, dir = "v") +
  labs(title = expression("Parameter estimation error vs speciation trend " * alpha[lambda])) +
  theme(panel.grid = element_blank(), strip.text = element_text(size = 10))
p

ggsave("Images/OBDD_NMSE_vs_alpha_lambda.pdf", p, width = 10, height = 5)
```

## NMSE vs net diversification trend $\alpha_\lambda - \alpha_\mu$

```{r fig.width=10, fig.height=5}
p <- ggplot(trend_data %>% filter(!par %in% c("αλ", "αμ")),
       aes(x = net_div_trend, y = NMSE)) +
  geom_point(alpha = 0.15, size = 1, color = "darkorange3") +
  geom_smooth(method = "loess", color = "darkorange3", linewidth = 0.5, se = TRUE,
              fill = "darkorange3") +
  scale_y_log10(name = "Normalized Mean Squared Error") +
  scale_x_continuous(name = expression("Net diversification trend (" * alpha[lambda] - alpha[mu] * ")")) +
  facet_wrap(~ ParametersExpr, scales = "fixed", nrow = 2, labeller = label_parsed, dir = "v") +
  labs(title = "Parameter estimation error vs net diversification trend") +
  theme(panel.grid = element_blank(), strip.text = element_text(size = 10))
p

ggsave("Images/OBDD_NMSE_vs_net_div_trend.pdf", p, width = 10, height = 5)
```

## NMSE vs extinction trend $\alpha_\mu$

```{r fig.width=10, fig.height=5}
p <- ggplot(trend_data %>% filter(!par %in% c("αλ", "αμ")),
       aes(x = true_αμ, y = NMSE)) +
  geom_point(alpha = 0.15, size = 1, color = "darkred") +
  geom_smooth(method = "loess", color = "darkred", linewidth = 0.5, se = TRUE,
              fill = "darkred") +
  scale_y_log10(name = "Normalized Mean Squared Error") +
  scale_x_continuous(name = expression("True " * alpha[mu] * " (extinction trend)")) +
  facet_wrap(~ ParametersExpr, scales = "fixed", nrow = 2, labeller = label_parsed, dir = "v") +
  labs(title = expression("Parameter estimation error vs extinction trend " * alpha[mu])) +
  theme(panel.grid = element_blank(), strip.text = element_text(size = 10))
p

ggsave("Images/OBDD_NMSE_vs_alpha_mu.pdf", p, width = 10, height = 5)
```

## Tree size as confound

The reviewer's hypothesis is that positive net-diversification trends produce larger
trees, which provide more information. We verify this expected confound.

```{r fig.width=10, fig.height=4}
trend_unique <- trend_data %>%
  filter(par == "λ") %>%
  select(dataset_nb, ntips, true_αλ, true_αμ, net_div_trend)

p1 <- ggplot(trend_unique, aes(x = true_αλ, y = ntips)) +
  geom_point(alpha = 0.3, size = 1.5) +
  geom_smooth(method = "loess", color = "steelblue") +
  scale_y_log10(name = "Number of tips") +
  labs(x = expression(alpha[lambda]),
       title = expression("Tree size vs " * alpha[lambda]))

p2 <- ggplot(trend_unique, aes(x = net_div_trend, y = ntips)) +
  geom_point(alpha = 0.3, size = 1.5) +
  geom_smooth(method = "loess", color = "steelblue") +
  scale_y_log10(name = "Number of tips") +
  labs(x = expression(alpha[lambda] - alpha[mu]),
       title = "Tree size vs net diversification trend")

p1 + p2

ggsave("Images/OBDD_tree_size_vs_trend.pdf", p1 + p2, width = 10, height = 4)
```
