---
title: "Calibration Plots"
output:
  pdf_document: 
    fig_height: 5.5
    fig_width: 10
    toc: yes
  html_notebook:
    code_folding: hide
  html_document:
    fig_width: 10
    toc: yes
---

Creation - jeremy.andreoletti@ens.fr - 05/06/2020

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r echo=FALSE}
library(ape)
library(phytools)
library(FossilSim)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(convenience)
```

# Single dataset
## Load files

MCMC trace 

```{r}
dataset_nb <- 3
#readLines(paste0("datasets_calibration_OBDD/dataset", dataset_nb, "/tree_full.tre"))
tree_full <- read.newick(paste0("/Volumes/data/INSANE_OBD/Calibration/datasets_calibration_OBDD/dataset", dataset_nb, "/tree_full.tre"))
tree_reconstructed <- read.newick(paste0("/Volumes/data/INSANE_OBD/Calibration/datasets_calibration_OBDD/dataset", dataset_nb, "/tree_reconstructed.tre"))

plotTree(tree_full)
if (!is.null(tree_full$node.label)){
  labelnodes(text=tree_full$node.label,node=1:tree_full$Nnode+Ntip(tree_full),interactive=FALSE,circle.exp=0.8,cex=0.7)
}
plotTree(tree_reconstructed, show.node.label=T)
if (!is.null(tree_reconstructed$node.label)){
  labelnodes(text=tree_reconstructed$node.label,node=1:tree_reconstructed$Nnode+Ntip(tree_reconstructed),interactive=FALSE,circle.exp=0.8,cex=0.7)
  }
```


```{r}
trace_file <- paste0("/Volumes/data/INSANE_OBD/Calibration/outputs/OBDD_SimulatedOBDD_dataset", dataset_nb, "_2000000iter.log")

trace <- read.table(trace_file, header=T)
names(trace) <- sub("lambda","λ",sub("mu","μ",sub("psi_","ψ",sub("omega_","ω",sub("sigma_","σ",sub("_root","",sub("alpha_","α",names(trace))))))))
head(trace)

#node_ages <- function(tree) return (paleotree::dateNodes(tree, rootAge = tree.max(tree)))
```

True values used in the simulations

```{r}
params_file <- paste("/Volumes/data/INSANE_OBD/Calibration/datasets_calibration_OBDD/dataset", dataset_nb, "/params.csv", sep="")
params <- read.table(params_file, header=T) %>% gather(par, value)
params <- params[1:10,]
params
```

## Compute the quantiles for the 100*alpha% credible interval

Example with $\alpha=0.95$

```{r}
alpha <- 0.95
quantile(trace$λ, probs = c(0.5-alpha/2, 0.5+alpha/2))
```

```{r fig.width=12, fig.asp=0.4}
hist(trace$λ, breaks = 50, border = alpha("darkgreen", 0.3), probability = T, col = alpha("lightgreen", 0.3),
     xlab = expression(λ), cex.main = 1.2, cex.lab = 1.5, 
     main = "Posterior distribution of lambda, the true value (red) is included in some credible intervals (green)")
for (alpha in 1:9/10) {
  lines(c(quantile(trace$λ, probs = c(0.5-alpha/2, 0.5+alpha/2))), c(1-alpha,1-alpha)*0.8, col="forestgreen", lwd=3)
  text(x = quantile(trace$λ, probs = 0.5+alpha/2)+0.2, y = (1-alpha)*0.8, labels = paste(alpha*100, "%"), cex=1.3, font=2)
  }
points(params[params$par=="λ",]$value, 0, col="red3", pch = 19, cex=2)
#invisible(dev.print(svg, "Images/Simulation_Based_Calibration_lambda.svg"))
```

```{r fig.width=12, fig.asp=0.4}
hist(trace$ω1, breaks = 50, border = alpha("darkblue", 0.3), probability = T, col = alpha("lightblue", 0.3),
     xlab = "ω1", cex.main = 1.2, cex.lab = 1.5, 
     main = "Posterior distribution of ω1, the true value (red) is included in some credible intervals (blue)")
for (alpha in 1:9/10) {
  lines(c(quantile(trace$ω1, probs = c(0.5-alpha/2, 0.5+alpha/2))), c(1-alpha,1-alpha)*0.15, col="blue3", lwd=3)
  text(x = quantile(trace$ω1, probs = 0.5+alpha/2)+1, y = (1-alpha)*0.15, labels = paste(alpha*100, "%"), cex=1.2, font=2)
  }
points(params[params$par=="ω1",]$value, 0, col="red3", pch = 19, cex=2)
#invisible(dev.print(svg, "Images/Simulation_Based_Calibration_psi.svg"))
```

Check for 20 $\alpha$ values and all parameters if the true value is included in the $100\alpha\%$ credible interval

```{r}
credible_interval <- function(par, alpha, trace){
  # Compute the interval range for a given alpha
  return (quantile(trace[,as.character(par)], probs = c(0.5-alpha/2, 0.5+alpha/2)))
}

for (alpha in 1:19/20) {
  credible_intervals <- sapply(params$par, credible_interval, alpha, trace)
  params[paste("alpha_", alpha, "_inf", sep="")] <- credible_intervals[1,]   # Inferior limit of the credible interval
  params[paste("alpha_", alpha, "_sup", sep="")] <- credible_intervals[2,]   # Superior limit of the credible interval
  # Check if the true values are included in the credible intervals
  params[paste("alpha_", alpha, "_in", sep="")] <- as.logical((params$value > params[paste("alpha_", alpha, "_inf", sep="")]) & 
                                                              (params$value < params[paste("alpha_", alpha, "_sup", sep="")]))
}
params
```

# All available datasets, OBDP

```{r}
# This line extracts dataset numbers from the filenames of log files in the "outputs" folder.
# The gsub function is used to remove the "COBD_SimulatedCOBD_dataset" and "_1000000iter.log" parts from the filenames.
# The list.files function lists all files in the "outputs" folder that match the pattern ".*1000000iter.log".
#dataset_nbs <- as.integer(gsub("OBDD_SimulatedOBDD_dataset|_2000000iter.log", "", list.files("outputs/OBDD", pattern="OBDD_SimulatedOBDD.*2000000iter.log")))
dataset_nbs <- as.integer(gsub("OBDD_SimulatedOBDD_dataset|_2000000iter.log", "", list.files("/Volumes/data/INSANE_OBD/Calibration/outputs", pattern="OBDD_SimulatedOBDD_dataset.*2000000iter.log")))

# These variables define the parameters for the analysis. They are set to their default values.
burnin_p <- 0              # Remove a given initial percentage of the MCMC trees
burnout_p <- 1             # Remove a given final percentage of the MCMC trees
rmtips <- 0                # Remove small trees with up to ... tips
subsampling_interval <- 1  # Subsample only one tree every each given interval in the trace to reduce autocorrelation

interval_inclusions <- c()   # Create an empty data frame to store the results
parameters_OBDD <- c()            # Create an empty data frame to store the results

for (dataset_nb in dataset_nbs) {

  # MCMC trace - Construct the path of the current log file based on the dataset number
  #trace_file <- paste0("outputs/OBDD/OBDD_SimulatedOBDD_dataset", dataset_nb, "_2000000iter.log")
  trace_file <- paste0("/Volumes/data/INSANE_OBD/Calibration/outputs/OBDD_SimulatedOBDD_dataset", dataset_nb, "_2000000iter.log")

  # Read the MCMC trace from the log file, skipping rows based on the subsampling interval
  trace <- read.table(trace_file, header=T)[c(T,rep(F, subsampling_interval-1)),]
  
  # Rename the columns of the trace data with their corresponding mathematical symbols
  names(trace) <- sub("lambda","λ",sub("mu","μ",sub("psi_","ψ",sub("omega_","ω",sub("sigma_","σ",sub("_root","",sub("alpha_","α",names(trace))))))))
  
  # Read the reconstructed tree from a Newick file based on the dataset number
  tree_reconstructed <- read.newick(paste0("datasets_calibration_OBDD/dataset", dataset_nb, "/tree_reconstructed.tre"))
  
  # Get the total number of tips and extant tips in the tree
  ntips <- Ntip(tree_reconstructed)
  ntipsextant <- length(getExtant(tree_reconstructed))
  nfossils <- sum(grepl("f", c(tree_reconstructed$tip.label, tree_reconstructed$node.label)))
  
  occurrences <- read.csv(paste0("datasets_calibration_OBDD/dataset", dataset_nb, "/occurrences.csv"), header = F)
  nocc <- nrow(occurrences)
  
  # Check if the number of tips in the tree is greater than the threshold for removing small trees
  if (ntips > rmtips) {
  
    # If burnin_p/burnout_p is non-zero, remove the corresponding percentage of initial/final MCMC trees
    if (burnin_p != 0) trace <- trace[-c(1:round(burnin_p*dim(trace)[1])),]
    if (burnout_p != 0) trace <- trace[-c(round(burnout_p*dim(trace)[1]):dim(trace)[1]),]
    
    # Read the true parameter values used in the simulations from a CSV file based on the dataset number
    # Convert the resulting data frame to a long format using the gather function from the dplyr package
    # Select the first 9 rows of the resulting data frame
    params_file <- paste("datasets_calibration_OBDD/dataset", dataset_nb, "/params.csv", sep="")
    params_OBDD <- read.table(params_file, header=T) %>% gather(par, value)
    params_OBDD <- params_OBDD[1:10,]
    
    # Create a data frame with information about the dataset number, parameter name, true value, and median value from the MCMC trace
    parameters_OBDD <- rbind(parameters_OBDD, data.frame(dataset_nb = rep(dataset_nb,nrow(params_OBDD)),
                                                         par = params_OBDD$par,
                                                         true_value = params_OBDD$value,
                                                         median_aPosteriori = sapply(trace[params_OBDD$par], median),
                                                         mean_aPosteriori = colMeans(trace[params_OBDD$par]),
                                                         var_aPosteriori = sapply(trace[params_OBDD$par], var),
                                                         MSE = colMeans(t(t(trace[params_OBDD$par]) - params_OBDD$value)**2),  # also equal to var_aPosteriori + (mean_aPosteriori - true_value)**2
                                                         NMSE = colMeans(t(t(trace[params_OBDD$par]) - params_OBDD$value)**2)/params_OBDD$value**2,
                                                         ntips = ntips,
                                                         ntipsextant = ntipsextant, 
                                                         nfossils = nfossils,
                                                         nocc = nocc,
                                                         row.names = NULL))
    
    # Loop over different alpha values in the range of 1/20 to 19/20
    for (alpha in 1:19/20) {
      
      # Compute the credible intervals for each parameter using the credible_interval function
      # The alpha value is passed as an argument to the function
      credible_intervals <- sapply(params_OBDD$par, credible_interval, alpha, trace)
      
      # Add columns to the params data frame representing the inferior limit and superior limit of the credible interval
      params_OBDD[paste("alpha_", alpha, "_inf", sep="")] <- credible_intervals[1,]
      params_OBDD[paste("alpha_", alpha, "_sup", sep="")] <- credible_intervals[2,]

      # Add a column to the params data frame indicating whether the true value falls within the credible interval
      params_OBDD[paste("alpha_", alpha, "_in", sep="")] <- as.logical((params_OBDD$value > params_OBDD[paste("alpha_", alpha, "_inf", sep="")]) & 
                                                                       (params_OBDD$value < params_OBDD[paste("alpha_", alpha, "_sup", sep="")]))
      
    }
    
    interval_inclusions <- rbind(interval_inclusions, data.frame(dataset_nb=rep(dataset_nb,nrow(params_OBDD)),params_OBDD[,c(1,grep("in$", names(params_OBDD)))]))
  }
}

parameters_OBDD <- parameters_OBDD[order(parameters_OBDD$dataset_nb),]
interval_inclusions <- interval_inclusions[order(interval_inclusions$dataset_nb),]
interval_inclusions
```

Compute the **truth recovery proportion** = percentage of datasets recovering the true value in their $100\alpha\%$ credible interval

```{r}
# Define a function to compute the truth recovery proportion given a parameter 'par'
truth_recovery_proportion <- function (par) {
  return(colMeans(interval_inclusions[interval_inclusions$par == par, c(-1, -2)]))
}

# Apply the truth recovery proportion function to each parameter in 'params_OBDD$par'
recovery <- data.frame(sapply(as.vector(params_OBDD$par), truth_recovery_proportion))

# Reorganize the 'recovery' dataframe for plotting
recovery_plot <- recovery %>% 
  rownames_to_column(var = "Alpha") %>% 
  pivot_longer(-c("Alpha"), names_to = "Parameters", values_to = "Proportions")

# Extract the 'Alpha' values from the row names and convert them to numeric
recovery_plot$Alpha <- as.double(gsub("alpha_|_in", "", recovery_plot$Alpha))

# Binomial proportion confidence interval
z <- 1.96
nb <- length(dataset_nbs)
recovery_plot$CI <- z*sqrt(recovery_plot$Proportions*(1-recovery_plot$Proportions)/nb)

# Wilson score interval https://en.wikipedia.org/wiki/Binomial_proportion_confidence_interval#Wilson_score_interval
recovery_plot$CorrectedProportions <- 1/(1+z^2/nb)*(recovery_plot$Proportions+z^2/(2*nb))
recovery_plot$CorrectedCI <- 1/(1+z^2/nb)*sqrt(recovery_plot$Proportions*(1-recovery_plot$Proportions)/nb + z^2/(4*nb^2))

recovery_plot
```

## Compare the empirical truth recovery proportion with the credible interval width, since the two are expected to correspond.

```{r}
# Create a mapping from your strings to their corresponding expressions
parameter_labels <- c("λ" = expression(lambda[0]),
                      "μ" = expression(mu[0]),
                      "αλ" = expression(alpha[lambda]),
                      "αμ" = expression(alpha[mu]),
                      "σλ" = expression(sigma[lambda]),
                      "σμ" = expression(sigma[mu]),
                      "ψ1" = expression(psi[1]),
                      "ψ2" = expression(psi[2]),
                      "ω1" = expression(omega[1]),
                      "ω2" = expression(omega[2]))

# Apply this mapping to the Parameters column in your data
recovery_plot$ParametersExpr <- factor(recovery_plot$Parameters, 
                                       levels = names(parameter_labels),
                                       labels = parameter_labels)
```

```{r fig.width=13, fig.asp = .16, message=FALSE}
p <- ggplot(recovery_plot, aes(x=Alpha, y=Proportions, color="Simulations")) +
  annotate("segment", x = 0, y = 0, xend = 1, yend=1, colour='gray35', lwd=0.2)+
  geom_point(alpha=0.7, colour='chartreuse4', cex=0.5) +
  geom_errorbar(aes(ymin=sapply(Proportions-CI, max, 0), ymax=sapply(Proportions+CI, min, 1)), width=.04, alpha=0.4, colour='chartreuse4')+
  scale_x_continuous(name = "Credible interval width", breaks = c(0,1)) +
  scale_y_continuous(name = "Truth recovery\nproportion", lim=c(0,1)) +
  facet_grid(. ~ ParametersExpr, labeller=label_parsed) +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1, size=8),
        axis.text.y = element_text(size=8),
        legend.position = "none")
p + ggtitle("Simulation-based evaluation of model calibration")

ggsave(paste0("Images/OBDD_truth_recovery_proportions", ifelse(rmtips>0, paste0("_rm", rmtips, "tip", ifelse(rmtips>1, "s", "")), ""), ".pdf"), p, width=9, height=1.5)
```

```{r fig.width=13, fig.asp = .16, message=FALSE}
#p <- ggplot(recovery_plot, aes(x=Alpha, y=Proportions, color="Simulations")) +
p <- ggplot(recovery_plot, aes(x=Alpha, y=CorrectedProportions, color="Simulations")) +
  annotate("segment", x = 0, y = 0, xend = 1, yend=1, colour='gray35', lwd=0.2)+
  geom_point(alpha=0.7, colour='chartreuse4', cex=0.5) +
  geom_errorbar(aes(ymin=CorrectedProportions-CorrectedCI, ymax=CorrectedProportions+CorrectedCI), width=.04, alpha=0.4, colour='chartreuse4')+
  scale_x_continuous(name = "Credible interval width", breaks = c(0,1)) +
  scale_y_continuous(name = "Truth recovery\nproportion", lim=c(0,1)) + 
  ggtitle("Simulation-based evaluation of model calibration") +
  facet_grid(. ~ ParametersExpr, labeller=label_parsed) +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1, size=8),
        axis.text.y = element_text(size=8),
        legend.position = "none")
p
```

## Compare the median a posteriori prediction to the true parameter values

```{r fig.width=12, fig.asp = .5, message=FALSE}
parameters_OBDD$ParametersExpr <- factor(parameters_OBDD$par, 
                                       levels = names(parameter_labels),
                                       labels = parameter_labels)

range_par <- sapply(unique(parameters_OBDD$par), function(P)range(parameters_OBDD[parameters_OBDD$par==P,]$median_aPosteriori, na.rm=T))
parameters_OBDD$par_min <- range_par[1,]
parameters_OBDD$par_max <- range_par[2,]

p <- ggplot(parameters_OBDD[!grepl("α", parameters_OBDD$par),], aes(x=true_value, y=median_aPosteriori, color="Simulations")) +
  geom_segment(aes(x=par_min, y=par_min, xend=par_max, yend=par_max, color="Perfect match"), alpha=0.7)+
  geom_point(alpha=0.2, size=2) +
  scale_x_sqrt(name = "True value used in simulation") +
  scale_y_sqrt(name = "Inferred median\na posteriori") + 
  scale_color_manual(name="Correspondence", values=c('gray35','darkblue'),
                     guide = guide_legend(override.aes = list(linetype = c("solid", "blank"), shape = c(NA, 16)))) +
  ggtitle("Simulation-based evaluation of model median predictions") +
  facet_wrap(. ~ ParametersExpr, scales="free", nrow=2, labeller=label_parsed) +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1, size=8),
        axis.text.y = element_text(size=8),
        legend.position = "none")
p
```
```{r fig.width=5, fig.asp = .6, message=FALSE}
p <- ggplot(parameters_OBDD[grepl("α", parameters_OBDD$par),], aes(x=true_value, y=median_aPosteriori, color="Simulations")) +
  geom_segment(aes(x=par_min, y=par_min, xend=par_max, yend=par_max, color="Perfect match"), alpha=0.7)+
  geom_point(alpha=0.2, size=2) +
  scale_x_continuous(name = "True value used in simulation") +
  scale_y_continuous(name = "Inferred median\na posteriori") + 
  scale_color_manual(name="Correspondence", values=c('gray35','darkblue'),
                     guide = guide_legend(override.aes = list(linetype = c("solid", "blank"), shape = c(NA, 16)))) +
  ggtitle("Accuracy of model median predictions") +
  facet_wrap(. ~ ParametersExpr, scales="free", nrow=1, labeller=label_parsed) +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1, size=8),
        axis.text.y = element_text(size=8),
        legend.position = "none")

p
```

## Compare the evolution of the prediction error with tree size

```{r fig.width=12, fig.asp = .5, message=FALSE}
parameters_OBDD$rel_error = parameters_OBDD$median_aPosteriori/parameters_OBDD$true_value

p <- ggplot(parameters_OBDD[!grepl("α", parameters_OBDD$par),], aes(x=ntips, y=rel_error, color="Simulations")) +
  annotate("segment", x = 2, xend = max(parameters_OBDD$ntips), y = 1, yend = 1, colour = "gray30", alpha=0.7, linewidth=1) +
  geom_point(alpha=0.2, size=2, colour = 'darkred') +
  scale_x_log10(name = "Number of tips", breaks=c(2,round(2.5**(2:6)/5)*5)) +
  scale_y_log10(name = "Estimation error\nMedian/True", breaks=signif(10**((-3:3)/3), 1)) + 
  ggtitle("Relative error in model median predictions for increasing tree size") +
  facet_wrap(. ~ ParametersExpr, scales = "free_y", nrow=2, labeller=label_parsed) +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1, size=8),
        axis.text.y = element_text(size=8),
        legend.position = "none")
p
```

```{r fig.width=12, fig.asp = .4, message=FALSE}
p <- ggplot(parameters_OBDD, aes(x=ntips, y=NMSE)) +
  #annotate("segment", x = 2, xend = max(parameters_OBDD$ntips), y = 0.01, yend = 0.01, colour = "gray30", alpha=0.7, linewidth=1) +
  geom_point(alpha=0.15, size=1, colour = 'darkred') +
  scale_x_log10(name = "Number of tips (extant and fossil)", breaks=c(2,round(2.5**(2:6)/5)*5)) +
  scale_y_log10(name = "Normalized Mean Squared Error", lim=c(0.002,1000)) + 
  geom_smooth(method = "glm", formula = 'y ~ x', method.args = list(family = gaussian(link = "identity")), color = alpha("darkred", 0.7), fill = alpha("darkred", 0.7), se = T) +
  facet_wrap(. ~ ParametersExpr, scales = "fixed", nrow=2, labeller=label_parsed, dir="v") +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1, size=8),
        axis.text.y = element_text(size=8),
        axis.title  = element_text(size=12),
        strip.text = element_text(size = 10),
        legend.position = "none")# + geom_abline(intercept=log10(0.002), slope=0) + geom_abline(intercept=log10(0.005), slope=0) + geom_abline(intercept=log10(500), slope=0) + geom_abline(intercept=log10(2000), slope=0)
p + ggtitle("Error in posterior estimates for increasing tree size")

ggsave(paste0("Images/OBDD_estimation_error_increasingTips", ifelse(rmtips>0, paste0("_rm", rmtips, "tip", ifelse(rmtips>1, "s", "")), ""), ".pdf"), p, width=9, height=9*0.4)
```

## Effect of occurrences

```{r fig.width=12, fig.asp = .5, message=FALSE}
min_ntips <- 0
parameters_OBDD_rmSmallTrees <- parameters_OBDD[parameters_OBDD$ntips>min_ntips,]

# Categorize datasets based on the number of fossil occurrences
fossil_nb_threshold <- quantile(parameters_OBDD_rmSmallTrees$nfossils, 1/2)
parameters_OBDD_rmSmallTrees$fossil_category <- ifelse(
  parameters_OBDD_rmSmallTrees$nfossils >= fossil_nb_threshold, 
  "Many fossils", 
  "Few fossils"
)

# Plot estimation error vs. number of fossil tips
p <- ggplot(parameters_OBDD_rmSmallTrees, aes(x = nocc, y = NMSE, color = fossil_category, fill = fossil_category)) +
  geom_point(alpha = 0.2, size = 2) +
  #geom_smooth(method = MASS::rlm, se = T) +
  geom_smooth(method = "glm", formula = 'y ~ x', se = T) +
  #annotate("segment", x = 10, xend = max(parameters_OBDD_rmSmallTrees$nocc), y = 0, yend = 0, colour = "gray30", alpha=0.7, linewidth=0.5, lty=2) +
  #scale_x_sqrt(name = "Number of fossil occurrences") +
  scale_x_log10(name = "Number of fossil occurrences") +
  scale_y_log10(name = "Normalized Mean Squared Error") +
  #coord_trans(y = scales::transform_exp(10)) +
  #scale_y_continuous(trans=scales::log1p_trans()) +
  #coord_trans(y = function()scales::new_transform("expm1", "expm1", "log1p")) +
  scale_color_manual(
    name = "Fossil category", 
    values = c('Few fossils' = 'deepskyblue3', 'Many fossils' = 'orange3')
  ) +
  scale_fill_manual(
    name = "Fossil category", 
    values = c('Few fossils' = 'deepskyblue3', 'Many fossils' = 'orange3')
  ) +
  ggtitle("Effect of Adding More Fossil Occurrences on Parameter Estimation Error") +
  facet_wrap(. ~ ParametersExpr, labeller = label_parsed, scales = "free", nrow = 2) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y = element_text(size = 8),
        strip.text = element_text(size = 12),
    legend.position = "bottom"
  )
p

# Save the plot
ggsave("Images/Estimation_error_vs_fossil_occurrences_OBDD.png", p, width = 12, height = 6)
```

```{r}
glm.NMSE <- lapply(unique(parameters_OBDD_rmSmallTrees$par), function(param){
  glm(log(NMSE) ~ log(ntipsextant) + sqrt(nfossils)*sqrt(nocc), 
      data = parameters_OBDD_rmSmallTrees[parameters_OBDD_rmSmallTrees$par == param,])
  })
names(glm.NMSE) <- unique(parameters_OBDD_rmSmallTrees$par)
lapply(glm.NMSE, summary)
#lapply(glm.NMSE, plot)
```

```{r}
# Initialize an empty list to store the summary tables for each parameter
glm_summaries <- list()

# Loop over each parameter in glm.NMSE
for (param in names(glm.NMSE)) {
  # Get the summary of the glm model
  glm_summary <- summary(glm.NMSE[[param]])
  
  # Extract the coefficients table
  coef_table <- as.data.frame(glm_summary$coefficients)
  
  coef_table$Parameter <- param
  coef_table$Predictor <- rownames(coef_table)
  coef_table <- coef_table[, c("Parameter", "Predictor", "Estimate", "Std. Error", "t value", "Pr(>|t|)")]
  glm_summaries[[param]] <- coef_table
}

# Combine all tables into a single data frame
glm_summary_table <- do.call(rbind, glm_summaries)
rownames(glm_summary_table) <- NULL

# View the summary table
as.data.frame(lapply(glm_summary_table, function(x)if (is.numeric(x)){formatC(x, format = "g", digits = 2)}else{x}))
```

```{r fig.width=8, fig.asp = .6, message=FALSE}
# Define the levels for Predictors and Parameters to ensure consistent ordering
glm_summary_table$Predictor <- factor(glm_summary_table$Predictor, levels = unique(glm_summary_table$Predictor))
glm_summary_table$Parameter <- factor(glm_summary_table$Parameter, levels = unique(glm_summary_table$Parameter))

# Predict the mean influence of each predictor by multiplying the coefficients by the average value of the predictor in the dataset
glm_summary_table$Mean_Predicted_Influence <- exp(glm_summary_table$Estimate*c(1, mean(log(parameters_OBDD_rmSmallTrees$ntipsextant)), mean(sqrt(parameters_OBDD_rmSmallTrees$nfossils)), mean(sqrt(parameters_OBDD_rmSmallTrees$nocc)), mean(sqrt(parameters_OBDD_rmSmallTrees$nfossils)*sqrt(parameters_OBDD_rmSmallTrees$nocc))))

glm_summary_table$ParametersExpr <- factor(glm_summary_table$Parameter, 
                                           levels = names(parameter_labels),
                                           labels = parameter_labels)

# Define custom breaks and colors
breaks <- c(log10(min(glm_summary_table$Mean_Predicted_Influence)),log10(1/3), -0.1, 0, 0.1, log10(3), log10(max(glm_summary_table$Mean_Predicted_Influence)))
colors <- c("darkgreen", "darkolivegreen3", "olivedrab1", "white", "lightcoral", "red3", "darkred")

p <- ggplot(glm_summary_table[glm_summary_table$Predictor != "(Intercept)",], aes(x = Predictor, y = ParametersExpr, fill = log10(Mean_Predicted_Influence), size = abs(log10(Mean_Predicted_Influence)), alpha = sapply(log10(`Pr(>|t|)`), function(x)max(x,-3)))) +
  geom_point(shape = 21, color = "black") +
  # Second layer: crosses for p-values > 0.05
  geom_point(data = subset(glm_summary_table, Predictor != "(Intercept)" & `Pr(>|t|)` > 0.01),
             shape = 4, alpha=0.4, stroke = 1) +
  scale_fill_gradientn(
    colors = colors,
    values = scales::rescale(breaks),
    name = "Average\nMultiplicative\nInfluence",
    breaks = log10(c(0.1, 1/3, 1, 3, 10)), 
    labels = c("1/10", "1/3", "1", "3", "10"),
    limits = round(c(min(breaks), max(breaks)))
  ) +
  scale_size_continuous(range = c(0, 10)) +
  scale_alpha_continuous(
    name = "P-value",
    limits = c(-1, -3),
    breaks = c(-1, -1, -2, -3), 
    labels = c(expression("">0.01), expression(""<=0.1), expression(""<=0.01), expression(""<=0.001)),
    trans = "reverse",
    guide = guide_legend(
      override.aes = list(size = c(5,5,5,5), shape = c(13,21,21,21), fill = "grey20")  # Custom points in the legend
    )
  ) +
  theme_minimal(base_size = 12) +
  scale_y_discrete(labels = scales::label_parse()) +
  labs(x = "Predictors", y = "Model Parameters") +
  theme(
    legend.position = "right",
    plot.title = element_text(size = 16, face = "bold"),
    axis.title.x = element_text(size = 14, vjust = -1),
    axis.title.y = element_text(size = 14, vjust = 3),
    axis.text.x = element_text(size=12, angle = 0, hjust = 0.5),
    axis.text.y = element_text(size=14)) +
  guides(size = "none")

p + ggtitle("Drivers of estimation error for OBDD parameters")

ggsave("Images/OBDD_drivers_estimation_error.pdf", p, width = 8, height = 8*0.6)
```

```{r}
# Create a data frame with predictors at average and baseline values
baseline_dataset <- data.frame(
  ntipsextant = mean(parameters_OBDD$ntipsextant),
  nfossils = 50,
  nocc = 0
)
baseline_dataset_noFossil <- data.frame(
  ntipsextant = mean(parameters_OBDD$ntipsextant),
  nfossils = 0,
  nocc = 0
)

plot_predicted_error <- function(dataset, occurrence_range, title){
  predicted_error <- sapply(glm.NMSE, predict, newdata = dataset, type = "response")
  
  predicted_error_over_baseline <- sapply(occurrence_range, function(nocc){
    new_dataset <- dataset
    new_dataset$nocc <- nocc
  
    # Predict expected value at baseline values
    predicted_error_new <- sapply(glm.NMSE, predict, newdata = new_dataset, type = "response")
    
    return (exp(predicted_error_new - predicted_error))
  })
  
  predicted_df <- as.data.frame(t(predicted_error_over_baseline))
  predicted_df$nocc <- occurrence_range
  predicted_long <- pivot_longer(predicted_df, cols = -nocc, names_to = "Parameter", values_to = "Predicted_Error_Ratio")
  
  # Plot the data
  ggplot(predicted_long, aes(x = nocc, y = Predicted_Error_Ratio, color = Parameter)) +
    geom_line() +
    labs(
      title = title,
      x = "Number of Occurrences (nocc)",
      y = "Predicted Error / Baseline Predicted Error"
    ) +
    ylim(0,2) +
    theme_minimal()
}

occurrence_range <- 0:150
plot_predicted_error(baseline_dataset_noFossil, occurrence_range, title ="Impact of including occurrences on the predicted estimation error (without fossils)")
plot_predicted_error(baseline_dataset, occurrence_range, title = "Impact of including occurrences on the predicted estimation error (with 50 fossils)")
```



```{r}
library(dplyr)
library(tidyr)

plot_predicted_error <- function(dataset, occurrence_range, title){
  # Compute predicted error and standard error at baseline
  predicted_error_list <- lapply(glm.NMSE, function(model){
    predict(model, newdata = dataset, type = "link", se.fit=TRUE)
  })
  
  # Initialize a list to store results
  results_list <- list()
  
  for (n in seq_along(occurrence_range)) {
    nocc <- occurrence_range[n]
    new_dataset <- dataset
    new_dataset$nocc <- nocc
    
    # Compute predicted error and standard error for new nocc
    predicted_error_new_list <- lapply(glm.NMSE, function(model){
      predict(model, newdata = new_dataset, type = "link", se.fit=TRUE)
    })
    
    # Compute ratios and confidence intervals
    results <- sapply(seq_along(glm.NMSE), function(i){
      fit0 <- predicted_error_list[[i]]$fit
      se0 <- predicted_error_list[[i]]$se.fit
      
      fit1 <- predicted_error_new_list[[i]]$fit
      se1 <- predicted_error_new_list[[i]]$se.fit
      
      delta_fit <- fit1 - fit0
      delta_se <- sqrt(se0^2 + se1^2)
      
      ratio <- exp(delta_fit)
      lower_CI <- exp(delta_fit - 1.96 * delta_se)
      upper_CI <- exp(delta_fit + 1.96 * delta_se)
      
      c(ratio=ratio, lower_CI=lower_CI, upper_CI=upper_CI)
    })
    
    # Convert results to data frame and set column names
    df <- data.frame(t(results))
    colnames(df) <- c("ratio", "lower_CI", "upper_CI")
    df$Parameter <- names(glm.NMSE)
    df$nocc <- nocc
    results_list[[n]] <- df
  }
  
  # Combine all data frames
  predicted_long <- do.call(rbind, results_list)
  
  # Plot the data with error bars
  ggplot(predicted_long, aes(x = nocc, y = ratio, color = Parameter, fill = Parameter)) +
    geom_line() +
    geom_ribbon(aes(ymin = lower_CI, ymax = upper_CI), alpha = 0.2, color = NA) +
    labs(
      title = title,
      x = "Number of Occurrences (nocc)",
      y = "Predicted Error / Baseline Predicted Error"
    ) +
    #ylim(0,2) +
    scale_y_log10() +
    theme_minimal()
}

occurrence_range <- 0:150
plot_predicted_error(baseline_dataset_noFossil, occurrence_range, title ="Impact of including occurrences on the predicted estimation error (without fossils)")
plot_predicted_error(baseline_dataset, occurrence_range, title = "Impact of including occurrences on the predicted estimation error (with 50 fossils)")
```

## Running times

```{r}
calculate_duration <- function(log_file) {
  # Read all lines from the log file
  log_lines <- readr::read_lines	(log_file)
  
  # Extract all lines that contain timestamps
  timestamp_lines <- grep("\\(\\d+\\.\\d+\\.\\d+\\) \\d{4}-\\d{2}-\\d{2} \\d{2}:\\d{2}:\\d{2}", log_lines, value = TRUE)
  
  # Extract the timestamps from these lines
  timestamps <- sub(".*\\((\\d+\\.\\d+\\.\\d+)\\) (\\d{4}-\\d{2}-\\d{2} \\d{2}:\\d{2}:\\d{2}).*", "\\2", timestamp_lines)
  
  # Convert the timestamps to POSIXct format
  time_values <- lubridate::ymd_hms	(timestamps)
  
  # Check if "Job terminated" is in the log file
  if (any(grepl("Job terminated", log_lines))) {
    # Extract the timestamp of the job termination
    end_time <- max(time_values)
  } else {
    # If "Job terminated" is not found, use the current time
    end_time <- Sys.time()
  }
  
  # Calculate the duration between the first timestamp and the end time in hours
  duration <- difftime(end_time, min(time_values), units = "days")
  
  return(as.numeric(duration))
}

durations <- c(sapply(0:99, function(dataset_nb)calculate_duration(paste0("/Volumes/data/INSANE_OBD/Calibration/Logs/OBDD.13112.", dataset_nb, ".log"))),
               sapply(0:399, function(dataset_nb)calculate_duration(paste0("/Volumes/data/INSANE_OBD/Calibration/Logs/OBDD.13118.", dataset_nb, ".log"))))
parameters_OBDD$running_time <- durations[parameters_OBDD$dataset_nb]
```

```{r fig.width=7, fig.asp = .7, message=FALSE}
# Plot estimation error vs. number of fossil tips
p <- ggplot(parameters_OBDD[parameters_OBDD$par == "λ",], aes(x = ntips, y = durations)) +
  geom_point(aes(colour = nocc), alpha = 0.5, size = 3) +
  #geom_smooth(method = "glm", formula = 'y ~ x', method.args = list(family = Gamma(link = "identity"))) +
  geom_smooth(method = "loess", formula = 'y ~ x', color="dodgerblue4", linewidth=0.5) +
  scale_x_continuous(name = "Number of tips") +
  scale_y_continuous(name = "Running time (days)") +
  labs(colour = "Number of\noccurrences") +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    axis.title  = element_text(size = 12),
    legend.background = element_rect(fill="transparent"),
    legend.position = c(0.9,0.25)
  )

p + ggtitle("Scaling of MCMC running time with tree size")

ggsave("Images/OBDD_running_time_vs_ntips.pdf", p, width = 5, height = 4)
```

```{r}
glm.running_time <- glm(running_time ~ ntipsextant + nfossils + nocc, data = parameters_OBDD[parameters_OBDD$par == "λ" & parameters_OBDD$ntips > 15,], family=Gamma(link = "identity"))

summary(glm.running_time)
plot(glm.running_time)
```

```{r}
library(mgcv)
gam.running_time <- gam(running_time ~ s(ntipsextant) + nfossils + nocc, data = parameters_OBDD[parameters_OBDD$par == "λ" & parameters_OBDD$ntips > 15,], family=Gamma(link = "identity"))

summary(gam.running_time)
plot(gam.running_time)
gam.check(gam.running_time)
```

```{r}
library(gamlss)

# Fit GAMLSS model with Gamma distribution
gamlss_model <- gamlss(
  running_time ~ pb(ntipsextant) + nfossils + nocc,
  sigma.fo = ~ ntipsextant,
  family = GA(mu.link = "identity", sigma.link ="log"),    # Gamma distribution
  data = subset(parameters_OBDD, par == "λ" & ntips > 15)
)

# Summary of the model
summary(gamlss_model)

# Diagnostic plots
par(mfrow = c(2, 2))
plot(gamlss_model)
```

# Effective sample sizes

```{r fig.width=10, fig.asp = .4}
outfiles <- list.files("/Volumes/data/INSANE_OBD/Calibration/outputs", pattern = "OBDD_SimulatedOBDD_dataset.*2000000iter.log", full.names=T)

runs <- lapply(outfiles, function(path){
  output <- list(trees = NA, ptable = read.table(path, header=T), gens.per.tree = 400)
  class(output) <- "rwty.chain"
  return (output)
  })

runs_ESS <- essContParam(runs, namesToExclude="iteration", tracer=T)
names(runs_ESS) <- paste0("ESS_dataset", sub("_2000000iter.log", "", sub("/Volumes/data/INSANE_OBD/Calibration/outputs/OBDD_SimulatedOBDD_dataset", "", outfiles)))

df <- rownames_to_column(data.frame(t(runs_ESS)), var = "Dataset")
df_long <- pivot_longer(df, cols = -Dataset, names_to = "Parameter", values_to="ESS")

p <- ggplot(df_long, aes(x=Parameter, y=ESS, fill=Parameter)) + geom_boxplot() + scale_y_log10()
p
```